The socialization of meritocracy and market justice preferences at school

### 2024-10-02 01-introduction.qmd
template mdpi
agregue un nuevo qmd para el template mdpi
RR-societies, abstract modification in paper-blinded.qmd and reviewed introduction.qmd

In the present paper, we address the role of the perception of meritocracy on the justification of social inequalities by eighth-grade students in Chile, a country characterized by a highly stratified educational system. In particular, we focus on justifying inequalities in health, education and pensions, which are traditional social policy areas where access to better services could be justified based on payment capacity, referred to as market justice [@lane_market_1986; @lindh_bringing_2023]. Most of the research in this realm has only considered adults, leaving aside the study of how beliefs in this field develop at student age as well as the impact of the school context and the family as the main socialization agencies [@batruch_belief_2022]. Regarding schools, the way in which they deal with unequal conditions of origin has been linked to the *hidden curriculum* [@chafel_schooling_1997], whereby students learn about distributive norms in society and mechanisms of justification of social differences.

>>>>

In the present paper, we address the role of the perception of meritocracy on the justification of social inequalities by eighth-grade students in Chile, a country characterized by a highly stratified educational system. In particular, we focus on justifying inequalities in health, education and pensions, which are traditional social policy areas where access to better services could be justified based on payment capacity, referred to as market justice [@lane_market_1986; @lindh_bringing_2023]. Market justice preferences refers to a belief in a market-based distributive principle, where goods and rewards are allocated according to individual merit, effort and contributions in a free-market economy (Lindh, 2015).

Most of the research on meritocracy and justice preferences has only considered adults, leaving aside the study of how beliefs in this field develop at student age as well as the impact of the school context and the family as the main socialization agencies [@batruch_belief_2022]. Studying beliefs about meritocracy and market justice preferences at school age is crucial because these early beliefs can shape individuals' attitudes towards inequality and social justice later in life. As Castillo (2019) argues, meritocratic ideals "serve as a principle that legitimizes the distribution of goods and rewards based on individual talent and effort," making it essential to understand how these beliefs develop during adolescence. Furthermore, early exposure to these beliefs may influence broader perceptions of fairness, as Reynolds (2014) highlights, "beliefs about the stratification system influence people's judgments about the fairness of inequality," suggesting that such attitudes formed at a young age can have lasting impacts. Additionally, García-Sierra (2023) emphasizes that the effects of holding meritocratic beliefs are not uniform, as these ideologies may reinforce socioeconomic disparities depending on students' social backgrounds.

### 2024-10-02 paper-blinded.qmd
RR-societies, abstract modification in paper-blinded.qmd and reviewed introduction.qmd

---
title: "The socialization of meritocracy and market justice preferences at school"
format:
  hikmah-pdf: 
    indent: false
    keep-tex: true
  hikmah-manuscript-pdf: 
    indent: false
    keep-tex: true
abstract: |
  Previous research has shown that schools justify student performance differences based on meritocratic ideals. However, the link between school meritocracy and students’ justification of inequalities has received little attention. This paper examines how school and family socialization influence students’ beliefs about meritocracy and market-based justifications for access to health, pensions, and education in Chile. Using data from the First Study of Civic Education in Chile, we employ a multilevel modeling strategy to analyze 5,047 students and parents nested within 231 schools. Our findings reveal that a significant proportion of Chilean students agree with market justice principles, at a higher rate than adults. Most students endorse meritocratic views, particularly the reward of effort. Multilevel estimations show strong associations between meritocratic beliefs and market justice preferences: students who believe in the reward of effort and talent are more likely to justify inequalities based on financial capacity. At the school level, market justice preferences are higher in high-status schools but lower in schools with better achievement. Additionally, the conditional influence of meritocracy beliefs weakens in schools with higher socioeconomic status and performance. These results suggest that meritocratic beliefs and market justice preferences begin early and are influenced by the school environment.
keywords: 
  - market justice 
  - meritocracy
  - socialization
  - family
  - schools
  - Chile
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/multidisciplinary-digital-publishing-institute.csl"
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm"
papersize: letter
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---
{{< pagebreak >}}

>>>>

---
title: "The socialization of meritocracy and market justice preferences at school"
format:
  hikmah-pdf: 
    indent: false
    keep-tex: true
  hikmah-manuscript-pdf: 
    indent: false
    keep-tex: true
abstract: |
  Previous research has shown that schools often justify student performance differences using meritocratic ideals. One potential consequence of such ideals is the legitimization of outcome inequalities across various spheres, including those traditionally associated with equality and redistribution. In this study, we argue that the promotion of meritocratic values during school-age can shape students' beliefs about meritocracy and influence their views on market-based access to health, pensions, and education. Using data from the 2017 National Study of Civic Education in Chile, which includes 5,047 8th-grade students from 231 schools, we estimated a series of multilevel models (lme4 library, R version 4.1.3) to test our hypotheses. Our findings show that a significant proportion of Chilean students agree with market justice principles—more so than adults. Most students endorse meritocratic views, particularly the notion that effort should be rewarded, which strongly correlates with market justice preferences: students who believe in meritocracy are more likely to justify inequalities based on financial capacity. At the school level, market justice preferences are higher in high-status schools but lower in schools with higher academic achievement. Furthermore, the conditional influence of meritocratic beliefs diminishes in schools with higher socioeconomic status and performance levels. These results suggest that the association between meritocratic beliefs and market justice preferences is already established at school age and is shaped by the school environment.
keywords: 
  - market justice 
  - meritocracy
  - socialization
  - family
  - schools
  - Chile
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/multidisciplinary-digital-publishing-institute.csl"
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm"
papersize: letter
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---
{{< pagebreak >}}
